To be FAIR: Theory Specification Needs an Update

### 2025-06-12 action_letter.Rmd
update 2025-06-12 16:23:17.396093

on behalf of all co-authors, I would like to express our sincere appreciation for your thoughtful comments. This is my (Caspar's) favorite review to date, and I especially value your perspective as a seasoned scholar/theoretician. All authors unanimously agreed that it would be a pity to deprive our readers of an opportunity to view FAIR theory through the lens of your lived experience - but we were not sure how to incorporate it. One author proposed to invite you as a co-author, others were concerned that this might constitute a conflict of interest.

We took the initiative to draft a paragraph that summarizes your comments on FAIR theory from the review, and mention your contribution in the Acknowledgements. We would be equally happy to credit your contribution with a co-authorship, depending on your - and the Editor's - preferences.

@@

To address this comment, we have removed as much of the specific detail as possible from the main text, and instead refer to the package vignettes (we have created an additional vignette to make this change possible, and extended several other vignettes).
There are already two worked examples for actual theories in the package vignettes,
which we now reference in the text.

>>>>

on behalf of all co-authors, I would like to express our sincere appreciation for your thoughtful comments. This is my (Caspar's) favorite review to date, and I especially value your perspective as a seasoned scholar/theoretician. All authors unanimously agreed that it would be a pity to deprive our readers of an opportunity to view FAIR theory through the lens of your lived experience - but we were not sure how to incorporate it. We invite your, and the Editor's, advice on how to best go about it.

As a starting point, we drafted a paragraph that summarizes your comments on FAIR theory from the review, and mentioned your contribution in the Acknowledgements. We are open to incorporating your thoughts and crediting your contribution in other ways as well, depending on what seems appropriate to you and the Editor.

@@

To address this comment, we have removed as much of the specific detail as possible from the main text, relegating it to the package vignettes instead.
To make this possible, we have substantially expanded the existing package vignettes and created one additional vignette.

There are now three worked examples for actual theories in the package vignettes:
[For Self-Determination Theory](https://cjvanlissa.github.io/theorytools/articles/formalizing_sdt.html), the [Dunning-Kruger Effect](https://cjvanlissa.github.io/theorytools/articles/dunning-kruger.html), and [Morris' Tripartite Model](https://cjvanlissa.github.io/theorytools/articles/causal-inference.html),
which we reference in the text.

### 2025-06-12 test.Rmd
update 2025-06-12 16:23:17.396093

Begin by expressing the theory as a structured digital object.
This can range from plain-text natural language statements to a more formal representation, such as a directed graph.
For example, the empirical cycle was originally described on page 28 of @degrootMethodologyFoundationsInference1969 as a series of natural language statements:

@@

However, we can also make an additional attempt to formalize the theory.
According to the levels of theory formalization,
the empirical cycle is presently defined at either the "theory" or "specification" level [@guestHowComputationalModeling2021].
We can further formalize the theory to the "implementation" level by specifying it in the DOT language for describing directed graphs^[Presented here in a simplified form; see the [tutorial](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html) for technical details].
This fulfills criterion I1 of the FAIR principles: using a formal language for knowledge representation [see Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv)).
Given the cyclical nature of the conceptual model, such an implementation might look like this:

@@

## 2. Create a Theory Archive

To meet the Interoperability and Reusability criteria,
it is important to properly document the theory file.
Firstly, add a README.md file with instructions for future users of your theory.
The `theorytools` package contains a [vignette](https://cjvanlissa.github.io/theorytools/articles/readme.html) on writing README files for theory.
Secondly, add a LICENSE file with the legal conditions for reuse.
We recommend explicitly waiving copyright with the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).
Place all files in a folder together - this folder becomes the theory archive.

## 3. Version Control the Archive

>>>>

Begin by creating an empty folder to hold all files associated with the theory - this folder will become the theory archive.
The first file to create is the theory itself.
This could be a plain-text file containing natural language statements,
or a more formal representation, such as a directed graph.
For example, the empirical cycle was originally described as a series of natural language statements [@degrootMethodologyFoundationsInference1969, p. 28]:

@@

Optionally, we can formalize the theory further.
According to a taxonomy of different levels of theory formalization,
the empirical cycle is presently defined at either the "theory" or "specification" level [@guestHowComputationalModeling2021].
To fulfill criterion I1 of the FAIR principles: using a formal language for knowledge representation [see Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv)),we can further formalize it to the "implementation" level by specifying it in the DOT language for describing directed graphs^[Presented here in a simplified form; see the [tutorial](https://cjvanlissa.github.io/theorytools/articles/fair-theory.html) for technical details].
Given the cyclical nature of the conceptual model, such an implementation might look like this:

@@

## 2. Document the Theory

To meet the Interoperability and Reusability criteria,
it is important to properly document the theory file.
Firstly, add a README.md file with instructions for future users of your theory.
The `theorytools` package contains a [vignette](https://cjvanlissa.github.io/theorytools/articles/readme.html) on writing README files for theory.
Secondly, add a LICENSE file with the legal conditions for reuse.
We recommend explicitly waiving copyright with the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).

## 3. Version Control the Theory Archive
